---
title: "Template based prediction of MS2 spectra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spectra-prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Template based prediction of lipid MS2 spectra

The <code>wormLipidBlastR</code> package allows the prediction of lipid spectra from their shorthand notation and template measurements. This enables to predict a large range of MS2 spectra for different lipids based on a few template measurements. The approach is inspired by the LipidBlast library from the Fiehn laboratory [1]. <code>wormLipidBlastR</code> utilizes functions from the <code>lipidomicsUtils</code> package for parsing of shorthand notations. Therefore this packages needs to be installed first from GitHub. Furthermore, the infrastructure for handling the predicted spectra requires functions from the <code>MSnbase</code> package.

```{r setup}
# installation of lipidomicsUtils
#devtools::install_github("michaelwitting/lipidomicsUtils")

library(wormLipidBlastR)
library(MSnbase)
```

# Prediction of spectra

After installation different functions are available for prediction. Currently supported lipid clases are PCs (1,2-diacyl-PCs, LM...), PEs (1,2-diacyl-PEs, LM...) and PS (1,2-diacyl-PSs, LM...). More lipid classes will be added soon. Predictions are based on masses for different building blocks as provived by the <code>lipidomicsUtils</code> package.

```{r}

```


[1] LipidBlast paper